Final analysis of Greenland narwhals trajectories with smoothSDE

### 2025-03-17 R/application/RCVM/fit_rcvm.Rmd
update application to narwhal data

library(smoothSDE)
library(ggplot2)
library(dplyr)
library(plotly)
library(htmlwidgets)
library(mgcv)
library(here)
library(xtable)
library(sf)
library(gganimate) # animated trajectories
library(doParallel)
library(foreach)
source(file.path(here("R","simulation_study_CRCVM","CVM_functions.R"))) 
```

@@

# Set the path to the directory containing the data
par_dir=here() #parent directory 
narwhal_data_path <- file.path(par_dir,"Data","preprocessed_data","narwhals")

@@

@@

@@

We try sde models where the parameters are smooth function of the covariates AngleNormal and Ishore.
Here, for simplicity we will always fix the measurement error to $30$ m. This choice is based on the paper of Weensven.
We will assess the sensitivity of our results to the hyperparameter later.

@@

#model formula
formulas <- list(mu1=~1,mu2=~1,tau =~s(theta_translated,k=10,bs="cc")+s(ID,bs="re"),nu=~s(ID,bs="re"))

#Fixed measurement error
sigma_obs=0.05
H=array(rep(sigma_obs^2*diag(2),n_post),dim=c(2,2,n_pre12))

#Fit baseline with measurement error estimated from the data
baseline_ctcrw0<- SDE$new(formulas = formulas,data = dataBE12,type = "CTCRW",
                          response = c("x","y"),
                          par0 = par0,
                          other_data=list("H"=H),
                          fixpar=c("mu1","mu2"))
baseline_ctcrw0$fit()

@@

get_raw_latex_tab<- function(sde,show=TRUE) {

@@

  if (show) {
    print(latex_table, type = "latex", include.rownames = FALSE)
  }

  return (latex_table)

@@

```{r estimates baseline_ctcrw0}
raw_tab_baseline_ctcrw0=get_raw_latex_tab(baseline_ctcrw0)
```

@@

get_final_baseline_tab<-function(sde,include_sigma_obs=FALSE,show=TRUE) {

@@

@@

  if (show) {
  print(
  xtab,type="latex",
  include.rownames = FALSE,
  sanitize.text.function = identity,  # Prevent escaping of LaTeX math symbols
  hline.after = c(-1, 0, nrow(table_data)),  
  table.placement="H")
  }

  return(xtab)

@@

final_tab_baseline_ctcrw0=get_final_baseline_tab(baseline_ctcrw0,include_sigma_obs=FALSE)

@@

```{r plot smooth parameters baseline_ctcrw0 }
baseline_ctcrw0$plot_par(var="theta_translated",
                         par_names=c("tau","nu"),n_post=100)

@@

## Baseline racvm with AngleNormal covariate in tau and omega

@@

# Fit baseline with measurement error estimated from the data
baseline_racvm0<- SDE$new(formulas = formulas,data = dataBE12,type = "RACVM_SSM",response = c("x","y"),
                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"))

@@

```{r estimates baseline_racvm0}
raw_tab_baseline_racvm0=get_raw_latex_tab(baseline_racvm0)
```

@@

final_tab_baseline_racvm0=get_final_baseline_tab(baseline_racvm0,include_sigma_obs=FALSE)

@@

## Baseline with AngleNormal and DistanceShore in omega and tau

@@

fomega=function(cov_data,a=1,b=2,D0=0.3,D1=1,sigma_D=0.3,sigma_theta=pi/6){
  Dshore=cov_data$DistanceShore
  theta=cov_data$theta
  if (is.null(Dshore)){
    Dshore=1/cov_data$Ishore
  }

  a*theta*(theta-pi/2)*(theta+pi/2)*exp(-Dshore/D0)/Dshore+
      b*(exp(-1/2*(((theta+pi/2/sqrt(3))/sigma_theta)^2+((Dshore-D1)/sigma_D)^2))-
           exp(-1/2*(((theta-pi/2/sqrt(3))/sigma_theta)^2+((Dshore-D1)/sigma_D)^2)))
}

@@

#parameters formulas
formulas <- list(mu1 = ~1 ,mu2 =~1,tau =~s(theta_translated,k=10,bs="cc")+s(ID,bs="re"),nu=~s(ID,bs="re"),
                   omega=~te(theta_translated,Ishore_translated,k=c(5,5),bs="cs"))

@@

```{r estimates baseline_racvm1}
raw_tab_baseline_racvm1=get_raw_latex_tab(baseline_racvm1)
```

@@

final_tab_baseline_racvm1=get_final_baseline_tab(baseline_racvm1,include_sigma_obs=FALSE)

@@

Ishore_values=c(1/0.25,1/2,1/10)
plots=lapply(1:length(Ishore_values),function(i){
  baseline_racvm1$plot_par(var="theta_translated",covs=data.frame("ID"="Asgeir","Ishore_translated"=Ishore_values[i]),
                         par_names=c("omega"),n_post=100)})
plots

@@

    # Check if the plot is a ggplot or plotly object
    if (inherits(plot, "ggplot")) {
      ggsave(file_path, plot, device = "pdf")
    } else if (inherits(plot, "plotly")) {
      saveWidget(plot, file_path)
  }
    })
```
Now, we would like to demonstrate how the Distance to shore influences the smooth parameter $\omega$.
We plot the estimated smooth parameter $\omega$ as a function of the angle for different values of the Distance to shore.

@@

estimate_parametric_omega=function(model,fixed_a=1) {

@@

    Dshore=seq(from=quantile(data$DistanceShore_translated,0.5),
               to=quantile(data$DistanceShore_translated,0.95),length.out=30)
    Ishore=1/Dshore
    theta_translated=seq(from=-pi,to=pi,length.out=30)
    grid<- as.data.frame(expand.grid(Ishore,theta_translated))
    colnames(grid) <- c("Ishore_translated","theta_translated")
    grid$ID="Asgeir"
    grid$Dshore_translated=1/grid$Ishore_translated

@@

    fit_optim <- optim(par = log(c(0.5,0.1,2,0.5,pi/6)),
                       fn = function(params) {

                        # Compute predicted z values
                        z_pred <- fomega(data_copy,fixed_a,exp(params[1]),
                                         exp(params[2]),exp(params[3]),exp(params[4]),exp(params[5]))

@@

    est_par=fit_optim$par

    print(exp(est_par))

    # Compute the predicted values using the estimated parameters
    z_pred_starting <- fomega(grid,fixed_a,exp(est_par[1]),
                              exp(est_par[2]),exp(est_par[3]),exp(est_par[4]),exp(est_par[5]))

@@

    return (list("est_par"=est_par,"splines_surface"=plot_splines,"parametric_surface"=plot_parametric))

@@

res_racvm1=estimate_parametric_omega(baseline_racvm1,fixed_a=2)

@@

#Fixed measurement error
sigma_obs=0.05
H=array(rep(sigma_obs^2*diag(2),length(all_data$time)),dim=c(2,2,length(all_data$time)))

#parameters formulas
formulas <- list(mu1 = ~1 ,mu2 =~1,tau =~s(theta_translated,k=10,bs="cc")+s(ID,bs="re"),nu=~s(ID,bs="re"),
                   omega=~te(theta_translated,Ishore_translated,k=c(5,5),bs="cs"))

@@

final_tab_all_racvm1=get_final_baseline_tab(all_racvm1,include_sigma_obs=FALSE)

@@

#parameters formulas
formulas <- list(mu1 = ~1 ,mu2 =~1,tau =~s(ID,bs="re"),nu=~s(ID,bs="re"),
                   omega=~te(theta_translated,Ishore_translated,k=c(5,5),bs=c("cs")))

# Fit baseline
baseline_racvm2<- SDE$new(formulas = formulas,data = dataBE12,type = "RACVM_SSM",
                    response = c("x","y"),par0 = par0,other_data=list("H"=H),
                    fixpar=c("mu1","mu2"))

@@

final_tab_baseline_racvm2=get_final_baseline_tab(baseline_racvm2,include_sigma_obs=FALSE)

@@

Ishore_values=c(1/0.25,1/2,1/5)
plots=lapply(1:length(Ishore_values),function(i){
  baseline_racvm2$plot_par(var="theta_translated",covs=data.frame("ID"="Asgeir","Ishore_translated"=Ishore_values[i]),
                         par_names=c("omega"),n_post=100)})
plots

@@

xmin=list("Ishore_translated"=1/3)
xmax=list("Ishore_translated"=1/0.3)
link=list("Ishore_translated"=\(x) 1/x)
xlabel=list("Ishore_translated"="Distance to shore")

@@

res_racvm2=estimate_parametric_omega(baseline_racvm2,fixed_a=2)

@@

## Simulate trajectories before exposure from fitted models

```{r,cache=TRUE}

@@

# Define time steps
delta=1/60/6
times_list=lapply(unique(dataBE12$ID),function(ID) {
  min=min(dataBE12[dataBE12$ID==ID,"time"])
  max=max(dataBE12[dataBE12$ID==ID,"time"])
  seq(min,max,by=delta)})
names(times_list)=unique(dataBE12$ID)
times_df <- data.frame(
  ID = rep(names(times_list), sapply(times_list, length)),
  time = unlist(times_list),row.names = NULL
)
n=length(times_df$time)
# Define covariates data (needed in the function simulate)
data_reg=data.frame("AngleNormal"=rep(0,n),"Ishore"=rep(1,n),"Ishore_translated"=rep(1,n),
                    "theta"=rep(0,n),"DistanceShore"=rep(1,n),"theta_translated"=rep(0,n))

@@

```{r}
est_par_racvm1=res_racvm1$est_par

formulas=list(tau=~s(theta_translated,k=10,bs="cc")+s(ID,bs="re"),nu=~s(ID,bs="re"),a=~1,b=~1,D0=~1,D1=~1,
              sigma_D=~1,sigma_theta=~1)
baseline_crcvm1<-SDE$new(formulas=formulas,data=dataBE12,type="CRCVM_SSM",response=c("x","y"),
               par0=c(baseline_racvm1$par()[1,c("tau","nu")],2,
                        exp(est_par_racvm1)))

baseline_crcvm1$update_coeff_re(baseline_racvm1$coeff_re()[1:20,])
baseline_crcvm1$update_lambda(baseline_racvm1$lambda()[1:3,])

baseline_crcvm1_trajectory=baseline_crcvm1$simulate(z0=z0_BE,data=data_reg,atw=atw,land=land,verbose=FALSE)

n=length(baseline_crcvm1_trajectory$time)

baseline_crcvm1_trajectory_sub=baseline_crcvm1_trajectory[seq(1,n,by=60),]

plot_baseline_crcvm1_trajectory=ggplot() +
  geom_sf(data=land$geometry,fill="grey")+
  coord_sf(datum=st_crs("+init=EPSG:32626 +units=km"))+
  geom_path(data=baseline_crcvm1_trajectory_sub, 
            aes(x = x, y = y,color=ID),size = 0.5, 
            lineend = "round",alpha=0.15) +     
  geom_point(data=baseline_crcvm1_trajectory_sub,
             aes(x = x, y = y,color=ID),size =0.5 ) +
  theme_minimal()
```

@@

set.seed(1)
est_par_racvm2=res_racvm2$est_par

formulas=list(tau=~s(ID,bs="re"),nu=~s(ID,bs="re"),a=~1,b=~1,D0=~1,D1=~1,
              sigma_D=~1,sigma_theta=~1)
baseline_crcvm2<-SDE$new(formulas=formulas,data=dataBE12,type="CRCVM_SSM",response=c("x","y"),
               par0=c(baseline_racvm2$par()[1,c("tau","nu")],2,
                        exp(est_par_racvm2)))

@@

baseline_crcvm2_trajectory=baseline_crcvm2$simulate(z0=z0_BE,data=data_reg,atw=atw,land=shrunk_land,verbose=FALSE)

@@

plot_baseline_crcvm2_trajectory=ggplot() +
  geom_sf(data=land$geometry,fill="grey")+
  coord_sf(datum=st_crs("+init=EPSG:32626 +units=km"))+
  geom_path(data=baseline_crcvm2_trajectory_sub, 
            aes(x = x, y = y,color=ID),size = 0.5, 
            lineend = "round",alpha=0.5) +     
  geom_point(data=baseline_crcvm2_trajectory_sub,
             aes(x = x, y = y,color=ID),size =0.5 ) +
  theme_minimal()

ggsave("plot_baseline_crcvm2_trajectory.pdf", plot=plot_baseline_crcvm2_trajectory,
       path="baseline_racvm2", width=10, height=6, units="in")

@@

baseline_crcvm1_trajectory_sub <- baseline_crcvm1_trajectory_sub %>% 
  mutate(data_type = "Simulated baseline_crcvm1")
baseline_crcvm2_trajectory_sub<- baseline_crcvm2_trajectory_sub %>% 
  mutate(data_type = "Simulated baseline_crcvm2")
observed_trajectory<- dataBE12%>% 
  mutate(data_type = "Observed")

baseline_trajectories <- bind_rows(baseline_crcvm1_trajectory_sub,
                                   baseline_crcvm2_trajectory_sub,observed_trajectory)

@@

@@

# add noise
noise=rmvn(length(baseline_crcvm2_trajectory_sub$time),rep(0,2),diag(rep(0.05^2,2)))
baseline_crcvm2_trajectory_sub_obs=baseline_crcvm2_trajectory[seq(1,n,by=30),c("time","x","y","ID")]
baseline_crcvm2_trajectory_sub_obs[,c("x","y")]=baseline_crcvm2_trajectory_sub_obs[,c("x","y")]+noise
baseline_crcvm2_trajectory_sub_obs=add_covs_parallel(baseline_crcvm2_trajectory_sub_obs)

@@

#initial parameters
par0 <- c(0,0,2,4,0)
init_lambda=c(exp(3),exp(3),m1$sp)

@@

#Fixed measurement error
sigma_obs=0.05
H=array(rep(sigma_obs^2*diag(2),length(baseline_crcvm2_trajectory_sub_obs$time)),
        dim=c(2,2,length(baseline_crcvm2_trajectory_sub_obs$time)))

@@

# Fit baseline
baseline_racvm2_check<- SDE$new(formulas = formulas,data = baseline_crcvm2_trajectory_sub_obs,type = "RACVM_SSM",
                    response = c("x","y"),par0 = par0,other_data=list("H"=H),
                    fixpar=c("mu1","mu2"))

@@

```
```{r}
get_final_baseline_tab(baseline_racvm2_check,include_sigma_obs=FALSE)
```

@@

@@

response0$fit()
```
```{r}
#plot parameters
xmin=list("AngleNormal"=-pi,"ExpShip"=1/45,"Ishore"=1/3)
xmax=list("AngleNormal"=pi,"ExpShip"=1/3,"Ishore"=1/0.1)
link=list("ExpShip"=(\(x) 1/x),"Ishore"=(\(x) 1/x))
xlabel=list("ExpShip"="Distance to ship","Ishore"="Distance to shore")
response0$get_all_plots(baseline=NULL,
      xmin=xmin, xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise")
```

@@

```{r fit response1}

@@

@@

@@

```{r display plots response3_sp}

@@

```{r results response_ctcrw_exp}

get_final_response_tab<-function(sde,show=TRUE) {

@@

```{r,results='asis',echo=FALSE}
final_tab_response_exp=get_final_response_tab(response_racvm_exp)

@@

```{r}
get_recovery_distances_tab<-function(p) {

@@

get_recovery_distances_tab(0.5)

>>>>

library(smoothSDE)
library(ggplot2)
library(ggpubr)
library(patchwork)
library(dplyr)
library(plotly)
library(htmlwidgets)
library(mgcv)
library(here)
library(xtable)
library(sf)
library(purrr)
library(gganimate) # animated trajectories
library(doParallel)
library(foreach)
source(file.path(here("R","simulation_study_CRCVM","CVM_functions.R"))) 
```

@@

# Set the path to the directory containing the data
par_dir=here() #parent directory 
narwhal_data_path <- file.path(par_dir,
                               "Data","preprocessed_data","narwhals")

@@

We bind data before exposure and after exposure in a single dataframe.

```{r bind data before and after exposure}

@@

# Interpolate every minute

```{r smooth and interpolate trajectories}
smooth_interpolate <- function(df) {

  # Create a sequence of regular time points (every minute)
  time_seq <- seq(min(df$time), max(df$time), by = 1/30)

  # Fit smoothing splines
  smooth_x <- smooth.spline(df$time, df$x)
  smooth_y <- smooth.spline(df$time, df$y)

  # Predict at regular intervals
  interp_x <- predict(smooth_x, as.numeric(time_seq))$y
  interp_y <- predict(smooth_y, as.numeric(time_seq))$y

  # Return the interpolated data
  data.frame(time = time_seq, x = interp_x, y = interp_y, ID = unique(df$ID))
}

# Apply the function for each ID
interpolated_dataBE12 <- dataBE12 %>%
  group_by(ID) %>%
  group_split() %>%
  map_dfr(smooth_interpolate)

# Plot the original and smoothed trajectories by ID
plot_interpolation=ggplot() +
  geom_point(data = dataBE12, aes(x = x, y = y, color = factor(ID)), size = 2) +
  geom_path(data = interpolated_dataBE12, aes(x = x, y = y, color = factor(ID))) +
  theme_minimal() +
  labs(color = "Trajectory ID") +
  ggtitle("GPS points and splines interpolation")
```

# Compute Dshore and theta from interpolation

                                   # Time steps
                                   dtimes <- sub_data[2:n_sub, "time"] -sub_data[1:(n_sub - 1), "time"]

  return(results)
}

interpolated_dataBE12=add_covs_parallel(interpolated_dataBE12)

@@

We try sde models where the parameters are smooth function 
of the covariates theta and Ishore.
Here, for simplicity we will always fix the measurement error to $50$ m. 
This choice is based on the paper of Weensven.
We will assess the sensitivity of our results to the hyperparameter later.

@@

#model formula
formulas <- list(mu1=~1,mu2=~1,tau =~s(theta,k=10,bs="cc")+s(ID,bs="re"),nu=~s(ID,bs="re"))

#Fixed measurement error
sigma_obs=0.01
H=array(rep(sigma_obs^2*diag(2),length(interpolated_dataBE12$time)),dim=c(2,2,length(interpolated_dataBE12$time)))

#Fit baseline with measurement error estimated from the data
baseline_ctcrw0<- SDE$new(formulas = formulas,data = interpolated_dataBE12,type = "CTCRW",
                          response = c("x","y"),
                          par0 = par0,
                          other_data=list("H"=H),
                          fixpar=c("mu1","mu2"))
baseline_ctcrw0$fit()

@@

make_raw_latex_tab<- function(sde) {

@@

  print(latex_table, type = "latex", include.rownames = FALSE)

@@

```{r raw estimates baseline_ctcrw0}
make_raw_latex_tab(baseline_ctcrw0)
```

@@

get_baseline_estimates<- function(sde,include_sigma_obs=FALSE) {

@@

  return(list("estimates"=mean_par,"std.error"=sd_par,"quantiles"=quant_par))

make_final_baseline_tab<-function(sde,include_sigma_obs=FALSE) {

  est=get_baseline_estimates(sde,include_sigma_obs)
  mean_par=est$estimates
  sd_par=est$std.error
  quant_par=est$quantiles
  parameter_names <- names(mean_par)
  estimates <- mean_par
  conf_intervals <- sapply(parameter_names, function(param) {
  sprintf("$[%.2f; %.2f]$", quant_par[[param]][1], 
          quant_par[[param]][2])
    })

@@

  print(
  xtab,type="latex",
  include.rownames = FALSE,
  sanitize.text.function = identity,  # Prevent escaping of LaTeX math symbols
  hline.after = c(-1, 0, nrow(table_data)),  
  table.placement="H")

@@

make_final_baseline_tab(baseline_ctcrw0,include_sigma_obs=FALSE)

@@

```{r plot smooth parameters baseline_ctcrw0 }
baseline_ctcrw0$plot_par(var="theta",
                         par_names=c("tau","nu"),n_post=100)

@@

## Baseline racvm with $\Theta$ covariate in tau and omega

@@

# Fit baseline with measurement error estimated from the data
baseline_racvm0<- SDE$new(formulas = formulas,data = dataBE12,
                          type = "RACVM_SSM",response = c("x","y"),
                    par0 = par0,other_data=list("H"=H),fixpar=c("mu1","mu2"))

@@

```{r raw estimates baseline_racvm0}
make_raw_latex_tab(baseline_racvm0)
```

@@

make_final_baseline_tab(baseline_racvm0,include_sigma_obs=FALSE)

@@

## Baseline with theta and DistanceShore in $\omega$ and $\tau$

@@

```{r get smooth penalty values}

fomega=function(cov_data,a=3,b=2,D0=0.4,D1=1,sigma_D=0.5,sigma_theta=pi/6){
  Dshore=cov_data$DistanceShore
  theta=cov_data$theta
  if (is.null(Dshore)){
    Dshore=1/cov_data$Ishore
  }

  a*theta*(theta-pi/2)*(theta+pi/2)*exp(-Dshore/D0)/Dshore+
      b*(exp(-1/2*(((theta+pi/2/sqrt(3))/sigma_theta)^2+
                     ((Dshore-D1)/sigma_D)^2))-
           exp(-1/2*(((theta-pi/2/sqrt(3))/sigma_theta)^2+
                       ((Dshore-D1)/sigma_D)^2)))
}

@@

#parameters formulas
formulas <- list(mu1 = ~1 ,mu2 =~1,
                 tau =~s(theta_translated,k=10,bs="cc")+s(ID,bs="re"),
                 nu=~s(ID,bs="re"),
                   omega=~te(theta_translated,Ishore_translated,k=c(5,5),bs="cs"))

@@

```{r raw estimates baseline_racvm1}
make_raw_latex_tab(baseline_racvm1)
```

@@

make_final_baseline_tab(baseline_racvm1,include_sigma_obs=FALSE)

@@

Ishore_values=c(1/0.25,1/2,1/10)
plots=lapply(1:length(Ishore_values),function(i){
  baseline_racvm1$plot_par(
    var="theta_translated",
    covs=data.frame("ID"="Asgeir","Ishore_translated"=Ishore_values[i]),
    par_names=c("omega"),n_post=100)})

plots

@@

Now, we would like to demonstrate how the Distance to shore 
influences the smooth parameter $\omega$.
We plot the estimated smooth parameter $\omega$ as a function 
of the angle for different values of the Distance to shore.

@@

```{r define smooth parametric estimate baseline_racvm1}

estimate_parametric_omega=function(model,fixed_a=NULL,probs=c(0.5,0.95)) {

@@

    Dshore=seq(from=quantile(data$DistanceShore_translated,probs[1]),
               to=quantile(data$DistanceShore_translated,probs[2]),length.out=30)
    Ishore=1/Dshore
    theta_translated=seq(from=-pi,to=pi,length.out=30)
    grid<- as.data.frame(expand.grid(Ishore,theta_translated))
    colnames(grid) <- c("Ishore_translated","theta_translated")
    grid$ID="Asgeir"
    grid$Dshore_translated=1/grid$Ishore_translated

@@

    if (is.null(fixed_a)) {

      fit_optim <- optim(par = log(c(1,0.5,0.1,2,0.5,pi/6)),
                       fn = function(params) {

                        # Compute predicted z values
                        z_pred <- fomega(data_copy,exp(params[1]),
                                         exp(params[2]),exp(params[3]),
                                         exp(params[4]),exp(params[5]),
                                         exp(params[6]))

@@

    } 
    else {
    fit_optim <- optim(par = log(c(0.5,0.1,2,0.5,pi/6)),
                       fn = function(params) {

                        # Compute predicted z values
                        z_pred <- fomega(data_copy,fixed_a,exp(params[1]),
                                         exp(params[2]),exp(params[3]),
                                         exp(params[4]),exp(params[5]))

    }
    est_par=fit_optim$par

    # Compute the predicted values using the estimated parameters
    if (is.null(fixed_a)) {

      z_pred_starting <- fomega(grid,exp(est_par[1]),
                              exp(est_par[2]),exp(est_par[3]),
                              exp(est_par[4]),exp(est_par[5]),
                              exp(est_par[6]))

    } else {
    z_pred_starting <- fomega(grid,fixed_a,
                              exp(est_par[1]),exp(est_par[2]),
                              exp(est_par[3]),exp(est_par[4]),
                              exp(est_par[5]))
    }

@@

    return (list("est_par"=exp(est_par),"splines_surface"=plot_splines,
                 "parametric_surface"=plot_parametric))

@@

res_racvm1=estimate_parametric_omega(baseline_racvm1,fixed_a=0.3)

@@

#Fixed measurement error
sigma_obs=0.05
H=array(rep(sigma_obs^2*diag(2),length(all_data$time)),
        dim=c(2,2,length(all_data$time)))

#parameters formulas
formulas <- list(mu1 = ~1 ,mu2 =~1,
                 tau =~s(theta_translated,k=10,bs="cc")+s(ID,bs="re"),nu=~s(ID,bs="re"),
                   omega=~te(theta_translated,Ishore_translated,k=c(5,5),bs="cs"))

@@

make_final_baseline_tab(all_racvm1,include_sigma_obs=FALSE)

@@

#Fixed measurement error
sigma_obs=0.01
H=array(rep(sigma_obs^2*diag(2),length(interpolated_dataBE12$time)),dim=c(2,2,length(interpolated_dataBE12$time)))

# Fit baseline
baseline_racvm2<- SDE$new(formulas = formulas,data = interpolated_dataBE12,
                    type = "RACVM_SSM",response = c("x","y"),par0 = par0,
                    other_data=list("H"=H),fixpar=c("mu1","mu2"))

@@

make_final_baseline_tab(baseline_racvm2,include_sigma_obs=FALSE)

@@

Ishore_values=c(1/0.25,1/2,1/5)
plots=lapply(1:length(Ishore_values),function(i){
  baseline_racvm2$plot_par(
    var="theta_translated",
    covs=data.frame("ID"="Asgeir","Ishore_translated"=Ishore_values[i]),
                         par_names=c("omega"),n_post=100)})
plots

@@

xmin=list("Ishore"=1/3)
xmax=list("Ishore"=1/0.1)
link=list("Ishore"=\(x) 1/x)
xlabel=list("Ishore"="Distance to shore")

@@

```{r get smooth parametric estimates baseline_racvm2}

res_racvm2=estimate_parametric_omega(baseline_racvm2,
                                     fixed_a=4,probs=c(0.75,0.95))

@@

To check the baseline model, we simulate trajectory using the estimated 
parameters, compare it to the observed data and use the simulated trajectory to 
estimate again the parameters. We check that the values estimated 
from the simulated trajectories are consistent with the values 
estimated from the data.

# Check baseline models

## Simulate trajectories using estimated parameters

```{r set up for simulation of trajectories,cache=TRUE}

@@

# Define time steps
delta=1/60/6
times_list=lapply(unique(dataBE12$ID),function(ID) {
  min=min(dataBE12[dataBE12$ID==ID,"time"])
  max=max(dataBE12[dataBE12$ID==ID,"time"])
  seq(min,max,by=delta)})
names(times_list)=unique(dataBE12$ID)
times_df <- data.frame(
  ID = rep(names(times_list), sapply(times_list, length)),
  time = unlist(times_list),row.names = NULL
)
n=length(times_df$time)
# Define covariates data (needed in the function simulate)
data_reg=data.frame("Ishore"=rep(1,n),"Ishore_translated"=rep(1,n),
                    "theta"=rep(0,n),"DistanceShore"=rep(1,n),
                    "theta_translated"=rep(0,n))

@@

@@

#set.seed(51)
set.seed(52)
est_par_racvm2=res_racvm2$est_par
fixed_a=4

formulas=list(tau=~s(ID,bs="re"),nu=~s(ID,bs="re"),a=~1,b=~1,D0=~1,D1=~1,
              sigma_D=~1,sigma_theta=~1)
baseline_crcvm2<-SDE$new(formulas=formulas,data=dataBE12,
                         type="CRCVM_SSM",response=c("x","y"),
               par0=c(baseline_racvm2$par()[1,c("tau","nu")],fixed_a,
                        est_par_racvm2))

@@

baseline_crcvm2_trajectory=baseline_crcvm2$simulate(
  z0=z0_BE,data=data_reg,atw=atw,land=shrunk_land,verbose=FALSE)

@@

plot_baseline_crcvm2_trajectory=ggplot() +
  geom_sf(data=shrunk_land$geometry,fill="grey")+
  coord_sf(datum=st_crs("+init=EPSG:32626 +units=km"))+
  geom_path(data=baseline_crcvm2_trajectory_sub, 
            aes(x = x, y = y,color=ID),size = 0.5, 
            lineend = "round",alpha=0.5) +     
  geom_point(data=baseline_crcvm2_trajectory_sub,
             aes(x = x, y = y,color=ID),size =0.5 ) +
  theme_minimal()+
  scale_color_viridis_d(labels=c("A1","A2","A3","A4","A5","A6"))+theme(
    axis.title = element_text(size = 18),
    axis.text = element_text(size = 14),
    legend.text = element_text(size = 18),
    axis.title.y = element_text(angle = 0)  # Rotate the y-axis label and adjust horizontal justification
  )

@@

baseline_crcvm2_trajectory_sub<- baseline_crcvm2_trajectory_sub %>% 
  mutate(data_type = "Simulated baseline_crcvm2")
observed_trajectory<- dataBE12%>% 
  mutate(data_type = "Observed")

baseline_trajectories <- bind_rows(baseline_crcvm2_trajectory_sub,
                                   observed_trajectory)

@@

                                   # Nearest points on shore
                                   sub_data <-cbind(sub_data,
                                                    nearest_boundary_points(as.matrix(sub_data[, c("x", "y")]), border))

@@

# add noise
baseline_crcvm2_trajectory_lf_obs=baseline_crcvm2_trajectory_sub[,c("time","x","y","ID")]
noise=rmvn(length(baseline_crcvm2_trajectory_lf_obs$time),rep(0,2),diag(rep(0.05^2,2)))

baseline_crcvm2_trajectory_lf_obs[,c("x","y")]=baseline_crcvm2_trajectory_lf_obs[,c("x","y")]+noise

baseline_crcvm2_trajectory_hf_obs=baseline_crcvm2_trajectory[seq(1,nrow(baseline_crcvm2_trajectory),by=30),c("time","x","y","ID")]
noise=rmvn(length(baseline_crcvm2_trajectory_hf_obs$time),rep(0,2),diag(rep(0.05^2,2)))
baseline_crcvm2_trajectory_hf_obs[,c("x","y")]=baseline_crcvm2_trajectory_hf_obs[,c("x","y")]+noise

baseline_crcvm2_trajectory_lf_obs=add_covs_parallel(baseline_crcvm2_trajectory_lf_obs)
baseline_crcvm2_trajectory_hf_obs=add_covs_parallel(baseline_crcvm2_trajectory_hf_obs)

@@

plot_baseline_crcvm2_trajectory_obs=ggplot() +
  geom_sf(data=shrunk_land$geometry,fill="grey")+
  coord_sf(datum=st_crs("+init=EPSG:32626 +units=km"))+
  geom_path(data=baseline_crcvm2_trajectory_lf_obs, 
            aes(x = x, y = y,color=ID),size = 0.2, 
            lineend = "round",alpha=0.5) +     
  geom_point(data=baseline_crcvm2_trajectory_lf_obs,
             aes(x = x, y = y,color=ID),size =0.2 ) +
  theme_minimal()+ scale_color_viridis_d(labels=c("A1","A2","A3","A4","A5","A6"))+theme(
    axis.title = element_text(size = 18),
    axis.text = element_text(size = 14),
    legend.text = element_text(size = 18),
    axis.title.y = element_text(angle = 0)  # Rotate the y-axis label and adjust horizontal justification
  )

plot_dataBE12=ggplot() +
  geom_sf(data=shrunk_land$geometry,fill="grey")+
  coord_sf(datum=st_crs("+init=EPSG:32626 +units=km"))+
  geom_path(data=dataBE12, 
            aes(x = x, y = y,color=ID),size = 0.2, 
            lineend = "round",alpha=0.5) +     
  geom_point(data=dataBE12,
             aes(x = x, y = y,color=ID),size =0.2 ) +
  theme_minimal()+ scale_color_viridis_d(labels=c("A1","A2","A3","A4","A5","A6"))+theme(
    axis.title = element_text(size = 18),
    axis.text = element_text(size = 14),
    legend.text = element_text(size = 18),
    axis.title.y = element_text(angle = 0)  # Rotate the y-axis label and adjust horizontal justification
  )

final_plot=ggarrange(plotlist=list(plot_dataBE12,plot_baseline_crcvm2_trajectory_obs),
                     ncol=2, nrow=1, common.legend = TRUE, legend="bottom",
                     labels= c("Observed","Simulated"))

ggsave("plot_baseline_crcvm2_trajectory.pdf", plot = final_plot,path="baseline_racvm2",width = 14, height = 8)
```

# Determine the common x-axis limits
x_min <- min(baseline_crcvm2_trajectory_hf_obs$DistanceShore, 
              dataBE12$DistanceShore_translated, na.rm = TRUE)
x_max <- max(baseline_crcvm2_trajectory_hf_obs$DistanceShore, 
              dataBE12$DistanceShore_translated, na.rm = TRUE)

# Plot the two histograms with consistent x-axis scales
Dshore_hist_crcvm2 <- ggplot(baseline_crcvm2_trajectory_hf_obs) +
  geom_histogram(aes(x = DistanceShore), color = "red", fill = "white", bins = 30) +
  theme_minimal() +
  xlim(x_min, x_max) +
  xlab("Distance to Shore") +
  ggtitle("Simulated data") +theme(
    axis.title = element_text(size = 18),
    axis.text = element_text(size = 14),
    legend.text = element_text(size = 18),
    axis.title.y = element_text(angle = 0)  # Rotate the y-axis label and adjust horizontal justification
  )

Dshore_hist_dataBE12 <- ggplot(dataBE12) +
  geom_histogram(aes(x = DistanceShore_translated), color = "red", fill = "white", bins = 30) +
  theme_minimal() +
  xlim(x_min, x_max) +
  xlab("Distance to Shore ") +
  ggtitle("Observed data")+theme(
    axis.title = element_text(size = 18),
    axis.text = element_text(size = 14),
    legend.text = element_text(size = 18),
    axis.title.y = element_text(angle = 0)  # Rotate the y-axis label and adjust horizontal justification
  )

# Combine the plots
Dshore_histograms <- Dshore_hist_crcvm2 + Dshore_hist_dataBE12

Dshore_histograms

# Save the plot as a PDF
ggsave("Dshore_histograms.pdf", plot = Dshore_histograms,path="baseline_racvm2", width = 10, height = 5)

#initial parameters
par0 <- c(0,0,1,1,0)
init_lambda=c(1/0.5^2,1/0.5^2,m1$sp)

@@

#Fixed measurement error
sigma_obs=0.05
H=array(rep(sigma_obs^2*diag(2),length(baseline_crcvm2_trajectory_hf_obs$time)),
        dim=c(2,2,length(baseline_crcvm2_trajectory_hf_obs$time)))

@@

# Fit baseline
baseline_racvm2_check<- SDE$new(formulas = formulas,
                                data = baseline_crcvm2_trajectory_hf_obs[baseline_crcvm2_trajectory_hf_obs$ID %in% c("Asgeir","Frederik","Kyrri","Nemo"),],type = "RACVM_SSM",
                    response = c("x","y"),par0 = par0,other_data=list("H"=H),
                    fixpar=c("mu1","mu2"))

@@

```{r}
make_final_baseline_tab(baseline_racvm2_check,include_sigma_obs=FALSE)
```

@@

```{r fit response all data}

@@

response0$fit()
```

```{r plot estimates response all data}
#plot parameters
xmin=list("AngleNormal"=-pi,"ExpShip"=1/45,"Ishore"=1/3)
xmax=list("AngleNormal"=pi,"ExpShip"=1/3,"Ishore"=1/0.1)
link=list("ExpShip"=(\(x) 1/x),"Ishore"=(\(x) 1/x))
xlabel=list("ExpShip"="Distance to ship","Ishore"="Distance to shore")
response0$get_all_plots(baseline=NULL,
      xmin=xmin, xmax=xmax,link=link,xlabel=xlabel,show_CI="pointwise")
```

@@

```{r fit response splines}

@@

```{r fit response log linear}

@@

```{r plot estimates response log linear}

@@

```{r display plots response log linear}

@@

```{r get final estimates response_racvm_exp}

make_final_response_tab<-function(sde,show=TRUE) {

@@

make_final_response_tab(response_racvm_exp)
```

@@

```{r}
make_recovery_distances_tab<-function(p) {

@@

make_recovery_distances_tab(0.5)
